A Survey of Physical Design Techniques of Information Systems

### 2010-07-27 cs848-db/Paper.tex

Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database system,
we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in
database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
performance. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed.
Data is stored in the form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and
access the data such that queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing
techniques, partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
techniques \cite{finkelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.

This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
are disk-based Relational Database Management Systems (RDBMS - traditional database systems), Main Memory Databases (MMDB), Data Warehouses, and XML
Databases. The aim of this survey
is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures and techniques have been
used in these
systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. Using the different
physical design methods used in disk based RDBMSs, we explore how these methods apply to the other systems and where the differences and similarities lie.

@@

Different variations of B-trees have been used. For example, B+trees are now the main method of indexing in current disk-based Relational Databases
\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the B+tree and the B-tree is that only the leaf nodes contain
data pointers in a B+tree, and leaf nodes contain pointers to each other as shown in Figure \ref{fig:b+tree}. In Prefix B+trees, instead of using the actual key
value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per node thus
decreasing the height of the tree. B+trees have been successful in relational databases since data is ultimately recorded in files. Since B+trees have a high
fanout, they allow less I/O operations to access a specific data record which is stored in a specific file.

Bitmap indices represent another type of index structures that is commonly used and cited in the literature. Using bitmaps as an index structure was first
implemented as part of the architecture of the Model 204 commercial database of Computer Corporation of America for the IBM 370 computer
\cite{patrick1987model}.
A bitmap index structure is based on a bitmap (i.e. bitset or bit array) data structure where individual bits correspond to values in unique records. Given a
table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap index defined for $C$ generates one bitmap vector for each of the
six different values ($B0-B5$). If the value of $C$ for record $x$ is set to 3, the corresponding bit in vector $B3$ is set to 1. Otherwise the bit is set to 0.

@@

Bitmap indices are best used when the set of possible key values in the index is small with a large number of records, e.g. gender attribute
\cite{oneil1995multi}. Bitmap indices outperform B+trees for low cardinality attributes. The reason is that given an attribute with low cardinality, defining a
B+tree index
on it means that the tree will have very few leaf nodes, each one storing a very long list of record identifiers (RIDs). On the other hand, defining a
bitmap index on that same attribute means that such long lists or RIDs will not be stored, rather bit vectors that represent the distinct attribute values will
be stored as previously explained. This leads to a direct benefit of consuming less storage space. Moreover, queries in that case can be answered using bit-wise
logical operations (e.g. and, or, not) which can improve query performance \cite{oneil1997improved}. Another advantage of bitmap indices is the ability to
perform more
efficient joins by creating bitmap join indices and using those indices to calculate the result of the join \cite{oneil1995multi}. Although there are many
benefits
gained from using bitmap indices, they poorly perform during updates. The reason is that when an update is to be executed to a bitmap index, the whole bit
vector
for the index is locked, blocking any access to the index for other operations. The lock is then released when the update is done. That is why bitmap indices
are usually used for read-mostly databases, a famous example of which are data warehouses (Section \ref{SEC-WAREHOUSES}).

@@

In general, indices are a way of organizing the logical structure of a database without altering its physical structure. However, sometimes it is desirable to
store related records near each other on disk for better locality. This can be achieved through implementing a clustered index on the desired key values.
Otherwise, the created index is known as a non-clustered index. The primary advantage of a clustered index is therefore the ordering of the physical data rows
in accordance with the index blocks that point to them. Therefore, fewer data block reads are required to answer a given query. The only caveat with using a
clustered index is that there can only be one clustered index per table because the physical organization of the table must be fixed
\cite{lightstone2007physical}.

@@

Partitioning in physical database design aims to reduct the workload on any one hardware component (E.g. an individual disk). This is achieved by partitioning
(dividing) the data over several disks \cite{lightstone2007physical}. There are two main categories of partitioning: horizontal partitioning and vertical
partitioning \cite{agrawal2004integrating}. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the
attributes of the table \cite{ceri1982horizontal, shin1985partitioning}. For example, dividing the data by date where all data dating less than year 2000 lies
in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width of the dataset by storing some
attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the size of the table. We will see in
Section~\ref{SEC-WAREHOUSES} that vertical partitioning is frequently used in Data Warehouses where not all the columns of a table are analyzed for decision
support.

@@

On the other hand, defining clustering keys, clustering scheme, and the granularity of clustering is very difficult. Accordingly, the problem of
designing an optimal clustering scheme has combinatorial complexity, having an extremely large search space \cite{jagadish1999snakes}.

@@

Garcia-Molina and Salem~\cite{garcia1992main} mention data access methods as one of the main challenges involved in MMDBs. Two considerations are taken when
designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
considerations taken for I/O operations are no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Cache
memories improve performance by holding recently referenced data~\cite{smith1982cache}. If the memory reference is found in the cache, then execution proceeds
at processor speed. Otherwise, the data has to be fetched from main memory. With the big difference between processor speed and main memory speed, cache misses
are still relatively expensive. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. Thus, a
cache miss is still relatively expensive in main memory databases.
Accordingly, we will see that many of the index data structures proposed for MMDBs focused on being cache conscious or cache sensitive. In
summary, a main memory index should be able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also
be noted that unlike disk based systems where it is advantageous to store actual attribute values in the index, main memory systems do not require that.
Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.

In this section, we talk about three families of index structures used in MMDB: B-trees, T-trees, and Binary Search Trees. There are more recent index
structures developed for MMDB such as J+trees~\cite{luan2009prefetching}, but which have still not gained much popularity, and so we do not discuss them in our
survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDBs. The T-tree is
the most used index for MMDBs (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.

 \begin{table}[t!]
\centering
\scalebox{0.8}{
\begin{tabular}{p{2.2cm}|p{6cm}|p{6cm}}
\hline
 \centering \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
\hline\hline
B-tree & 
\raggedright -Good storage utilization \linebreak  \linebreak
\raggedright -Reasonably quick searching \linebreak \linebreak
-Fast updating 
& 
-Reasonable cache behavior if node fits in cache line\\
\hline
B+tree &
\raggedright -Reasonably quick searching \linebreak \linebreak
\raggedright -Fast updating
&
-Reasonable cache behavior if node fits in cache line\\
\hline
CSB+tree & 
-Same features of B+ Trees \linebreak \linebreak
-Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
 &
-Good cache performance due to improved locality \& lower tree height (because of larger node size)\\
\hline

pB+tree & 
-Same features of B+ Trees
             &
-Improves cache miss performance by prefetching more data in each cache miss \linebreak \linebreak
-Node size is also increased leading to the same benefits of CSB+ Trees\\
\hline
T-tree &  
-Contains pointers to data values instead of the values themselves leading to better space utilization \linebreak \linebreak
-Many sorted keys per node \linebreak \linebreak
-Only minimum and maximum keys used for comparisons \linebreak \linebreak
\raggedright -Contains record pointers in every key \linebreak \linebreak
-Good update \& storage characteristics if Btrees.
             &
-Cache behavior was not considered at time of design \linebreak \linebreak
-Record pointers waste space in cache\\
\hline
CST & 
-Binary search tree for maximum value of each node leads to faster search \linebreak \linebreak
-More space efficient by removing pointers \& using array storage. Index calculation used to allocate values &
-Node size is aligned with cache line size to avoid misses \linebreak \linebreak
-Higher usage of cached data in binary search tree\\
\hline
CSS-trees & 
-Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node) \linebreak \linebreak
-Mainly suitable for read environments &
-Good cache behavior since $m$ is chosen to fit in the cache line size\\

@@

\subsubsection*{B-trees}
Although B-trees, explained in the previous section, are originally designed for disk based database systems, they were also found to be useful in
MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in MMDBs, the B-tree would be
more suitable from a storage perspective since there is no gain from keeping all the data in the leaves. In a main memory system, this
would only waste space without improving performance. On the other hand, the B+ Tree uses multiple keys to search within a node. This means that
if the node fits in a cache line, this cache load can satisfy more than more comparison leading to better cache utilization~\cite{rao1999cache}. Additionally,
less data per node is stored in a B+tree which means there is better cache utilization. Therefore, B+trees in general have reasonable cache behavior, and
variations of them have been adapted in MMDBs.

@@

Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+trees. They call the new index structure prefetching B+tree
(pB+tree). Since current computer systems can prefetch different data simultaneously, the authors take advantage of this. The idea here is to increase the node
size of B+ Trees and employ prefectching at the same time. In a B+tree, every time we move down a level in the tree, a cache miss occurs. Accordingly,
prefetching larger sizes nodes means that we can overlap multiple cache misses before they occur. This is because in one cache miss, extra data is retrieved in
parallel with the originally requested data if it can be predicted sufficiently early. Figure~\ref{fig:prefetching} shows how prefetching retrieves the same
data in less time. This method, however, does not eliminate the full cache miss latency with every level in the tree.

\begin{figure}[!t]
\centering
\includegraphics[width=7cm]{figs/prefetching.png}
\caption{Performance of various B+ Tree searches where a cache miss to memory takes 150 cycles and a subsequent access can begin 10 cycles
later~\cite{chen2001improving}}
\label{fig:prefetching}
\end{figure}

\subsubsection*{T Trees}
Lehmen and Carey~\cite{lehman1986study} introduce a new index structure, the T-tree, which combines the features from AVL Trees and B-trees that are suited to
main memory. Figure~\ref{fig:ttree} shows the proposed structure of the T-tree. They ran some experiments on data residing in main memory to compare the
performance of the T-tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The
structures compared included  AVL Trees, simple arrays, B-trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their
experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T-trees gave the best performance for a mix
of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics. The T-tree
was the first index structure specifically designed for main memory databases. The T-tree has many sorted keys per node, and only uses the minimum and maximum
keys for comparison. Additionally, every key contains a record pointer.

@@

Binary Search Trees have poor cache behavior when the array is much bigger than the cache. Rao and Ross~\cite{rao1999cache} propose the Cache Sensitive Search
Tree (CSS-tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard code the traversal within a node
such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree where $m$ is the number of keys per node.
However, CSS-trees are only suitable for read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.

\subsubsection{Materialized Views}
From our literature survey, it seems that materialized views are not frequently used in MMDB. This makes sense since the problems materialized views try to
address problems are not highly present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing
the data in memory. Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit. However, this does not mean that they
cannot be used at all in MMDBs. As with all other physical design decisions, it rather depend on the query workload and the relative benefits of implementing
the technique. Accordingly, if the data is very large and may benefit from using materialized views, there is no reason not to use them.

@@

Partitioning is used in disk based systems to divide the data into the physical pages in a way that will yield the best performance. This could be through
vertical partitioning or horizontal partitioning, as previously explained. In main memory databases, we do not have the issue of expensive disk access which
makes partitioning in memory unnecessary as it will not save any costs. However, partitioning is needed in the secondary storage of the data on
disk~\cite{gruenwald1990database}. Disk storage is still used as a backup of main memory databases. When a crash occurs, a reload of the database from disk or
archive memory takes place. In order to avoid page faults caused by referencing data that is still being reloaded, the reloading process must be designed to be
efficient in loading the important data such that the database users are not affected. An efficient MMDB reload takes into consideration the number of I/Os for
reload and the number of main memory references during transaction processing. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing} show that
horizontal partitioning is most suitable when the database performs more modifications than tuple deletions and more selections that projections and joins.
Vertical partitioning is chosen in the other cases. They also show that if we just concentrate on the reload performance (i.e the number of I/Os for
reload), then vertical partitioning is always the best choice.

@@

XML (eXtensible Markup Language) is increasingly being used as a data exchange format. Accordingly, the ability to store and manage XML documents is needed.
Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents and their parts, maintained by a system
having capabilities to manage and control the collection itself and the information represented by that collection.'' Salminena and Tompa also highlight some of
the requirements of an XML database system. These requirements include querying parts of an XML document, and
transformations of XML documents. The main challenge in XML databases is path navigation since it is a very expensive operation.

@@

In this section, we will discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries
and updates are faced in XML database design, but these are beyond the scope of this survey.

@@

For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, B+trees are still used to index the XML documents.
However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique
identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
Four index files are created for XML data where all the indexes are based on B+-trees. One index manages the collection hierarchy, one index collects nodes
in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
occurrences and is used by the full text search extensions. Oracle uses a similar technique for indexing its XML content where an XMLIndex table is created for
XML entries~\cite{oracleXML}. This table contains a unique identifier for the XPath leading to the node, the row id of the table used to store the XML data,an
order key that identifies the hierarchal positioning of the node, a locater key used for fragment extraction, and the text value of the node.

TIMBER also uses a similar numbering schema. There are minor differences between the two numbering techniques, but the main idea is the same. The numbering is
based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
different reasoning to produce its numbering schema. Sedna assigns string identifiers to nodes such that they are lexically ordered according to the position
of the node. Despite the differences in the ways the numbering is constructed, it seems that existing indexing techniques can be adapted for XML databases. The
challenge is how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
reflect the XML document structure. Additionally, in native XML databases, documents are usually grouped into collections, and an index may be used to search
within these collections. This is different from the node index which searches within a document.

@@

The main idea behind using materialized views is to rewrite queries using these views so that they run more efficiently. This is very important in XML
databases since XQuery and XPath queries are complicated~\cite{arion2007structured}. Accordingly, materialized views are used in XML databases to improve
performance. The semi-structured nature of
XML data, and the expressiveness of most XML querying languages, view selection for optimization becomes a more complicated
problem~\cite{tang2009materialized}. Therefore, the choice of which views to materialize becomes more complicated in XML database. Much work has been
done on rewriting XML queries using views~\cite{arion2007structured, balmin2004framework, aouiche2006clustering, tang2008multiple}. However, this problem is not
within the scope this survey.

>>>>

Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database system,
we usually think of which data we need to keep track of and how we will properly organize it into meaningful tables. However, this is not the only phase in
database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
performance. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed.
Data is stored in the form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and
access the data such that queries are answered and executed efficiently. In the traditional relational database system, this involves selecting proper indexing
techniques, partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
techniques \cite{finkelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.

This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
are disk-based Relational Database Management Systems (RDBMS - traditional database systems), Main Memory Databases (MMDB), Data Warehouses, and XML
Databases. The aim of this survey
is not to explain the details of every data structure used in physical design, but rather to compare how different data structures and techniques have been
used in these
systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. Using the different
physical design methods used in disk based RDBMSs, we explore how these methods apply to the other systems and where the differences and similarities lie.

@@

Different variations of B-trees have been used. For example, B+trees are now the main method of indexing in current disk-based Relational Databases
\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the B+tree and the B-tree is that only leaf nodes contain
data pointers in a B+tree, and leaf nodes contain pointers to each other as shown in Figure \ref{fig:b+tree}. In Prefix B+trees, instead of using the actual key
value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per node thus
decreasing the height of the tree. B+trees have been successful in relational databases since data is ultimately recorded in files. Since B+trees have a high
fanout, they allow less I/O operations to access a specific data record which is stored in a specific file.

Bitmap indices represent another type of index structures that is commonly used and cited in the literature. Using bitmaps as an index structure was first
implemented as part of the architecture of the Model 204 commercial database of Computer Corporation of America for the IBM 370 computer
\cite{patrick1987model}.
A bitmap index structure is based on a bitmap (i.e. bitset or bit array) data structure where individual bits correspond to values in unique records. Given a
table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap index defined for $C$ generates one bitmap vector for each of the
six different values ($B0-B5$). If the value of $C$ for record $x$ is set to value 3, the corresponding bit in vector $B3$ is set to 1. Otherwise the bit is set
to 0.

@@

Bitmap indices are best used when the set of possible key values in the index is small with a large number of records, e.g. gender attribute
\cite{oneil1995multi}. Bitmap indices outperform B+trees for low cardinality attributes. The reason is that given an attribute with low cardinality, defining a
B+tree index
on it means that the tree will have very few leaf nodes, each one storing a very long list of record identifiers (RIDs). On the other hand, defining a
bitmap index on that same attribute means that such long lists or RIDs will not be stored, rather bit vectors that represent the distinct attribute values will
be stored as previously explained. This leads to a direct benefit of consuming less storage space. Moreover, queries in that case can be answered using bit-wise
logical operations (e.g. and, or, not) which can improve query performance \cite{oneil1997improved}. Another advantage of bitmap indices is the ability to
perform more
efficient joins by creating bitmap join indices and using those indices to calculate the result of the join \cite{oneil1995multi}. Although there are many
benefits
gained from using bitmap indices, they poorly perform during updates. The reason is that when an update is to be executed on a bitmap index, the whole bit
vector
for the index is locked, blocking any access to the index for other operations. The lock is then released when the update is done. That is why bitmap indices
are usually used for read-mostly databases, a famous example of which are data warehouses (Section \ref{SEC-WAREHOUSES}).

@@

Sometimes it is desirable to store related records near each other on disk for better locality. This can be achieved through implementing a clustered index on
the desired key values.
Otherwise, the created index is known as a non-clustered index. The primary advantage of a clustered index is therefore the ordering of the physical data rows
in accordance with the index blocks that point to them. Therefore, fewer data block reads are required to answer a given query. The only caveat of using a
clustered index is that there can only be one clustered index per table because the physical organization of the table must be fixed
\cite{lightstone2007physical}.

@@

Partitioning in physical database design aims to reduce the workload on any one hardware component (E.g. an individual disk). This is achieved by partitioning
(dividing) the data over several disks \cite{lightstone2007physical}. There are two main categories of partitioning: horizontal partitioning and vertical
partitioning \cite{agrawal2004integrating}. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the
attributes of the table \cite{ceri1982horizontal, shin1985partitioning}. For example, dividing the data by date where all data dating less than year 2000 lies
in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width of the dataset by storing some
attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the size of the table. We will see in
Section~\ref{SEC-WAREHOUSES} that vertical partitioning is frequently used in Data Warehouses where not all the columns of a table are analyzed for decision
support.

@@

On the other hand, defining clustering keys, clustering scheme, and the granularity of clustering is very difficult. Accordingly, the problem of
designing an optimal clustering scheme has combinatorial complexity leading to an extremely large search space \cite{jagadish1999snakes}.

@@

Garcia-Molina and Salem~\cite{garcia1992main} mention data access methods as one of the main challenges involved in MMDBs. Two considerations are taken when
designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
considerations taken for I/O operations are no longer applicable. Instead of focusing on disk access and disk storage, the data structures for main memory
databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having cache conscious index structures. Cache
memories improve performance by holding recently referenced data~\cite{smith1982cache}. If the memory reference is found in the cache, then execution proceeds
at processor speed. Otherwise, the data has to be fetched from main memory. With the big difference between processor speed and main memory speed, cache misses
are still relatively expensive. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. Thus, a
cache miss is still relatively expensive in main memory databases.
Accordingly, we will see that many of the index data structures proposed for MMDBs focused on being cache conscious or cache sensitive. In
summary, a main memory index should be able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also
be noted that unlike disk-based systems where it is advantageous to store actual attribute values in the index, main memory systems do not require that.
Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.

\begin{figure}[!t]
\centering
\includegraphics[width=7cm]{figs/prefetching.png}
\caption{Performance of various B+tree searches where a cache miss to memory takes 150 cycles and a subsequent access can begin 10 cycles
later~\cite{chen2001improving}}
\label{fig:prefetching}
\end{figure}

In this section, we talk about three families of index structures used in MMDB: B-trees, T-trees, and Binary Search Trees. There are more recent index
structures developed for MMDB such as J+trees~\cite{luan2009prefetching}, but which have still not gained much popularity, and so we do not discuss them in our
survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDBs. The T-tree is
the most used index for MMDBs (E.g. MySQL CLuster). However, there are other indices used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.

 \begin{table}[t!]
\centering
\scalebox{0.8}{
\begin{tabular}{p{2.2cm}||p{6cm}|p{6cm}}
 \centering \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
\hline\hline
B-tree & 
\raggedright -Good storage utilization \linebreak  \linebreak
\raggedright -Reasonably quick searching \linebreak \linebreak
-Fast updating 
& 
-Reasonable cache behavior if node fits in cache line\\
\hline
B+tree &
\raggedright -Reasonably quick searching \linebreak \linebreak
\raggedright -Fast updating
&
-Reasonable cache behavior if node fits in cache line\\
\hline
CSB+tree & 
-Same features of B+ Trees \linebreak \linebreak
-Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
 &
-Good cache performance due to improved locality \& lower tree height (because of larger node size)\\
\hline

pB+tree & 
-Same features of B+ Trees
             &
-Improves cache miss performance by prefetching more data in each cache miss \linebreak \linebreak
-Node size is also increased leading to the same benefits of CSB+trees\\
\hline
T-tree &  
-Contains pointers to data values instead of the values themselves leading to better space utilization \linebreak \linebreak
-Many sorted keys per node \linebreak \linebreak
-Only minimum and maximum keys used for comparisons \linebreak \linebreak
\raggedright -Contains record pointers in every key \linebreak \linebreak
-Good update \& storage characteristics if B-trees
             &
-Cache behavior was not considered at time of design \linebreak \linebreak
-Record pointers waste space in cache\\
\hline
CST-tree & 
-Binary search tree for maximum value of each node leads to faster search \linebreak \linebreak
-More space efficient by removing pointers \& using array storage. Index calculation used to allocate values &
-Node size is aligned with cache line size to avoid misses \linebreak \linebreak
-Higher usage of cached data in binary search tree\\
\hline
CSS-tree & 
-Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node) \linebreak \linebreak
-Mainly suitable for read environments &
-Good cache behavior since $m$ is chosen to fit in the cache line size\\

@@

\subsubsection*{B-trees}
Although B-trees, explained in the previous section, are originally designed for disk based database systems, they were also found to be useful in
MMDB~\cite{lehman1986study}. In disk based systems, the B+tree is more used than the regular B-tree. However, in MMDBs, the B-tree would be
more suitable from a storage perspective since there is no gain from keeping all the data in the leaves. In a main memory system, this
would only waste space without improving performance. On the other hand, the B+tree uses multiple keys to search within a node. This means that
if the node fits in a cache line, this cache load can satisfy more than more comparison leading to better cache utilization~\cite{rao1999cache}. Additionally,
less data per node is stored in a B+tree which means there is better cache utilization. Therefore, B+trees in general have reasonable cache behavior, and
variations of them have been adapted in MMDBs.

@@

Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+trees. They call the new index structure prefetching B+tree
(pB+tree). Since current computer systems can prefetch different data simultaneously, the authors take advantage of this. The idea here is to increase the node
size of B+trees and employ prefetching at the same time. In a B+tree, every time we move down a level in the tree, a cache miss occurs. Accordingly,
prefetching larger sizes nodes means that we can overlap multiple cache misses before they occur. This is because in one cache miss, extra data is retrieved in
parallel with the originally requested data if it can be predicted sufficiently early. Figure~\ref{fig:prefetching} shows how prefetching retrieves the same
data in less time. This method, however, does not eliminate the full cache miss latency with every level in the tree.

\subsubsection*{T-trees}
Lehmen and Carey~\cite{lehman1986study} introduce a new index structure, the T-tree, which combines the features from AVL Trees and B-trees that are suited to
main memory. Figure~\ref{fig:ttree} shows the proposed structure of the T-tree. They ran some experiments on data residing in main memory to compare the
performance of the T-tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The
structures compared included  AVL Trees, simple arrays, B-trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their
experiments showed that for unordered data, Modified Linear Hashing gives the best performance, and for ordered data, T-trees give the best performance for a
mix
of searches, inserts and deletes. This is mainly because a T-tree node contains many elements which results in good update and storage characteristics. The
T-tree
was the first index structure specifically designed for main memory databases. The T-tree has many sorted keys per node, and only uses the minimum and maximum
keys for comparison. Additionally, every key contains a record pointer.

@@

Binary Search Trees have poor cache behavior when the array is much bigger than the cache. Rao and Ross~\cite{rao1999cache} propose the Cache Sensitive Search
Tree (CSS-tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard code the traversal within a node
such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree where $m$ is the number of keys per node.
However, CSS-trees are only suitable for read environments since they eliminate nearly all child pointers, thus removing the support for incremental updates.

\subsubsection{Materialized Views}
From our literature survey, it seems that materialized views are not frequently used in MMDB. This makes sense since the problems materialized views try to
address are not highly present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing
the data in memory. Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit. However, this does not mean that they
cannot be used at all in MMDBs. As with all other physical design decisions, it rather depend on the query workload and the relative benefits of implementing
the technique. Accordingly, if the data is very large and may benefit from using materialized views, there is no reason not to use them.

@@

Partitioning is used in disk-based systems to divide the data into the physical pages in a way that will yield the best performance. This could be through
vertical partitioning or horizontal partitioning, as previously explained. MMDBs do not have the issue of expensive disk access which
makes partitioning in memory unnecessary as it will not save any costs. However, partitioning is needed in the secondary storage of the data on
disk~\cite{gruenwald1990database}. Disk storage is still used as a backup of main memory databases. When a crash occurs, a reload of the database from disk or
archive memory takes place. In order to avoid page faults caused by referencing data that is still being reloaded, the reloading process must be designed to be
efficient in loading the important data such that the database users are not affected. An efficient MMDB reload takes into consideration the number of I/Os for
reload and the number of main memory references during transaction processing. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing} show that
horizontal partitioning is most suitable when the database performs more modifications than tuple deletions and more selections than projections and joins.
Vertical partitioning is suitable for the other cases. They also show that if we just concentrate on the reload performance (i.e the number of I/Os for
reload), then vertical partitioning is always the best choice.

@@

XML (eXtensible Markup Language) is being increasingly used as a data exchange format. Accordingly, the ability to store and manage XML documents is
needed.
Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents and their parts, maintained by a system
having capabilities to manage and control the collection itself and the information represented by that collection.'' Salminena and Tompa also highlight some of
the requirements of an XML database system. These requirements include querying parts of an XML document, and
transformations of XML documents. The main challenge in XML databases is path navigation since it is a very expensive operation.

@@

In this section, we will discuss how the physical design elements need to be adapted in both cases. Other challenges such as expressing and executing queries
and updates are faced in XML database design, but these are beyond the scope of this survey.

@@

For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, B+trees are still used to index the XML documents.
However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique
identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
Four index files are created for XML data where all the indexes are based on B+trees. One index manages the collection hierarchy, one index collects nodes
in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
occurrences and is used by the full text search extensions. Oracle uses a similar technique for indexing its XML content where an XMLIndex table is created for
XML entries~\cite{oracleXML}. This table contains a unique identifier for the XPath leading to the node, the row id of the table used to store the XML data,an
order key that identifies the hierarchal positioning of the node, a locater key used for fragment extraction, and the text value of the node.

TIMBER~\cite{jagadish2002timber} also uses a similar numbering schema. There are minor differences between the two numbering techniques, but the main idea is
the same. The numbering is
based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
different reasoning to produce its numbering schema. Sedna assigns string identifiers to nodes such that they are lexically ordered according to the position
of the node. Despite the differences in the ways the numbering is constructed, it seems that existing indexing techniques can be adapted for XML databases. The
challenge is how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
reflect the XML document structure. Additionally, in native XML databases, documents are usually grouped into collections, and an index may be used to search
within these collections. This is different from the node index which searches within a document.

@@

The main idea behind using materialized views is to rewrite queries using these views so that they run more efficiently. This is very important in XML
databases since XQuery and XPath queries are complicated~\cite{arion2007structured}. Accordingly, materialized views are used in XML databases to improve
performance. The semi-structured nature of
XML data, and the expressiveness of most XML querying languages makes view selection for optimization a more complicated
problem~\cite{tang2009materialized}. Therefore, the choice of which views to materialize becomes more complicated in XML database. Much work has been
done on rewriting XML queries using views~\cite{arion2007structured, balmin2004framework, aouiche2006clustering, tang2008multiple}. However, this problem is not
within the scope this survey.
